A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes

### 2023-11-29 pilot-publication/manuscript.Rmd
Update manuscript

**Introduction** Several trauma life support programs aim to improve trauma outcomes but there is no evidence from controlled trials to show the effect of these programs on patient outcomes. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The patient recruitment rate was X and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r stringr::str_to_sentence(english::english(n.m30d.atls))` (`r p.m30d.atls`%) patients died within 30 days in the ATLS arm, `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response, causes 4.5 millions deaths every year
[@GBD2018]. Almost 10% of the global burden of disease is due
to trauma and trauma is the top contributor to the burden of disease in
children and adults aged 10 to 49 years [@Vos2020].

@@

The proprietary Advanced Trauma Life Support (ATLS) is the most
established trauma life support training program and more than one
million doctors in over 80 countries have been trained in the program
[@acsAtls2018]. Uptake in low- and middle income countries (LMIC)
has been slow, potentially due to high costs [@Kadhum2020].

The free Primary Trauma Care (PTC) program is the most widely spread
alternative program. The goal of PTC is to improve trauma care in LMIC
[@ptcfPtc2018]. Like ATLS, doctors in over 80 countries have been
trained in PTC, and the program has been endorsed by the World Health
Organization (WHO), among other international organizations
including several professional societies [@ptcfPtc2018].

Despite the widespread use of these training programs there are no
controlled trials showing that they impact patient outcomes
[@Mohammad2013; @Jayaraman2014; @Kadhum2020]. A systematic review
report that hospital trauma life support training is associated with a
reduction in patient mortality, based on studies that were either
observational or quasi-experimental without a control group
[@Jin2021].<!-- Note that some of the studies that Jin include state
that there was no effect of ATLS -->

We will perform a pilot study that aims to assess the feasibility of
conducting a cluster randomised controlled trial comparing ATLS and
PTC with standard care. Recent methodological guidelines indicate that
the design of efficient cluster randomised controlled trials requires
data on probable or target effect sizes, proportion of participants
with the outcome (if binary), and the intracluster correlation
coefficient [@Hemming2017]. The objectives of this pilot study were
to:

- Estimate probable effect sizes on patient outcomes associated with
  ATLS and PTC compared with standard care, estimate the proportion of
  participants with the outcome (if binary), and estimate the
  intracluster correlation coefficient, as a basis for future sample
  size calculations.

@@

## Trial Design
We piloted a three-armed cluster randomised controlled trial. There
were two intervention arms, ATLS and PTC training, and one control
arm, standard care. We collected data for four months in all three
arms, first during a one month observation phase and then during a
three month intervention phase (or continued observation in the
control arm). This design allowed us to assess outcomes both as
final values and as change from baseline.

@@

## Study Setting
We conducted this pilot in seven Indian tertiary hospitals, where
neither ATLS, PTC, nor any other trauma life support training program
is routinely taught. The seven hospitals were Seth G. S. Medical
College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
College and General Hospital, Mumbai; HBT Medical College And Dr. R N
Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
Medical Education & Research, Chandigarh.

@@

### Resident Doctor Participants
Resident doctors doing their speciality training in surgery or
emergency medicine managing trauma patients in the emergency
department and who were expected to remain in the participating
hospitals for at least one year. To facilitate administration each
surgical department is divided into units, which manages the out
patient department, emergency department, operating rooms etc on
different days each week. One or two, out of typically six, units'
residents were selected from each hospital. One unit consists of at
least three faculty and three to twelve residents.

To be eligible, units had to have a maximum of 25% of the doctors
trained in either ATLS, PTC, or similar training programs before the
start of the pilot. Those residents who had received training in the
last five years were considered as trained. The figure of 25% was
decided through consensus in the research team, to balance feasibility
and contamination of results.

Consent was sought from the residents in each of the intervention
groups before they underwent the ATLS or PTC training. The consent
form for residents is available as Supplemental Material 2. We did
not ask for consent from residents at the units in the control
hospitals as their practice was not affected by this pilot and we
did not collect any personal identifiable data on them.

@@

## Interventions
In each intervention arm one or two units', out of typically six,
residents per hospital providing emergency care to trauma patients
were trained in either ATLS or PTC. For the purpose of this pilot
study, we will target to train a minimum of 75% of residents in each
unit. If residents dropped out or changed units after training but
before data collection was completed we planned to conduct additional
training if needed to meet the 75% criterion, but this was not
required. We did not train the units' faculty, as they are typically
not involved in the initial management of trauma patients.

The ATLS training was conducted in an ATLS certified training centre
in Mumbai, according to the standard ATLS curriculum
[@acsAtls2018]. The PTC training was conducted in New Delhi, according
to the standard PTC curriculum [@ptcfPtc2018]. These courses were
conducted over a period of 2.5 to 3 days. The residents certified
"pass" were considered as trained in respective courses.

The control group provides standard care with no intervention.

### Modifications
Both ATLS and PTC are standard training programs with fixed curricula
[@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or
content of these programs during this pilot.

### Adherence
The intervention was the training in either ATLS or PTC and resident
participants were required to adhere to, i.e. participate in, the
training, to be eligible for passing. We did not consider adherence to
training contents during care delivery as adherence to the trial
intervention.

@@

Although training in an advanced trauma life support course is part of
the curriculum it is optional and not doing this training does not
result in failure to obtain post graduation completion.

@@

The primary participant outcome was all cause mortality within 30 days
from the time of arrival to the emergency department. The primary
outcome and most secondary outcome were assessed and compared both as
final values and as change from baseline. All outcomes that pertain to
the individual participant level are detailed in Supplemental Material
3. We decided to include a large number of outcomes,
including some more exploratory, so that we can test their feasibility
and relevance.

@@

### Patients
Patients were screened for eligibility as they arrive at the emergency
department. Eligible patients were approached in the emergency
department to consent to follow up, if they are conscious. If they
were unconscious a patient representative was approached to consent to
follow up. Once the patient is conscious we approached the patient to
affirm the patient representative's consent. We followed up patients
at discharge, at 24 hours after arrival at the emergency department,
and at 30 days after arrival at the emergency department.

### Residents
Participating units were be screened for eligibility once hospitals
confirmed their participation. All residents in eligible units were
approached to consent to training if their hospital was randomised to
either of the intervention arms. Training was conducted approximately
one month after the study started.

@@

## Patient and public involvement
In this study, we conducted community consultations to collect
inputs from patients, their caregivers, patient groups, and resident
doctors to be used in the selection of outcome measures and
implementation of the full-scale trial; following the Guidance for
Reporting Involvement of Patients and the Public (GRIIPS 2)
[@Staniszewska2017].

During the pilot study, interviews were conducted with post-discharge
trauma patients and their caregivers to identify the outcomes most
relevant to them. These patients were identified through the medical
registers of the participating hospitals, contacted through telephone,
and after receiving their consent they were interviewed as per their
convenience. Their consent form is available as Supplemental
Material 5.  Additionally, members from non-government organizations
working with trauma patients and the hospital Social Service Section
were contacted for their views on contextual patient-centred outcomes
for trauma patients. Their consent form is available as Supplemental
Material 6. For feasibility, these interviews were held in each of the
cities where the participating centres are located. The commonest
patient-centred outcomes reported across all the locations will be
incorporated into the evaluation of the effects of the different
training programs and standardized care on patient outcomes.

Similarly, the inputs of resident doctor participants at each
participating centre were collected during the pilot study. A
discussion with the residents were conducted to document any
challenges or suggestions they may have in the scheduling or
implementation of the training programs. These inputs will be
incorporated in the final study.

A summary of the findings of the study as well as their inputs will be
shared with those who participated in the interviews and
discussions. A meeting will be held with the patient participants, at
each city, where the changes in the measured patient-centered outcomes
would be presented to them. Another meeting will be held with the
resident doctors at each hospital to present the confidence of the
residents after being trained.  Any suggestions and reflections from
the participants during the meetings will be used as inputs for
planning the final study.

## Data management
We used the online data collection tool KoboToolbox [@KoboToolbox],
accessible only over a virtual private network (VPN), for each
participating hospital to upload pseudonymised data to secure
servers. Data validation techniques such as restricted values or
values within a specific range were used to avoid ambiguous data
entries and ensure the validity of the data. Ambiguous responses and
errors were resolved after discussion with the core team during weekly
meetings. An instruction manual or codebook for data variables was
prepared to ensure consistency in data entry. This manual was referred
to during the project data collection and variable descriptions were
visible for each variable in the online data collection
tool. Pseudonymised data were stored at the centralised server. The
data were only accessible by the project's principal investigator or
by delegation by the project principal investigator only.

@@

# Competing Interests
Several authors are Advanced Trauma Life Support or Primary Trauma
Care instructors.

@@

```{r secondary-outcomes, echo = FALSE, results = "asis"}
table1::t1kable(secondary.outcomes.table, longtable = TRUE) %>%
    kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8)

```

>>>>

**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The patient recruitment rate was `r recruitment.rate.patients`% and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r stringr::str_to_sentence(english::english(n.m30d.atls))` (`r p.m30d.atls`%) patients died within 30 days in the ATLS arm, `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response, causes 4.3 millions deaths every year
[@injuries2020]. Almost 10% of the global burden of disease is due
to trauma and trauma is the top contributor to the burden of disease in
children and adults aged 10 to 49 years [@Vos2020].

@@

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) is the most
established trauma life support training program and more than one
million doctors in over 80 countries have been trained in the programme
[@acsAtls2018]. Uptake in low- and middle income countries (LMIC)
has been slow, potentially due to high costs [@Kadhum2020].

The free Primary Trauma Care (PTC) program is the most widely spread
alternative programme. The goal of PTC is to improve trauma care in LMIC
[@ptcfPtc2018]. Like ATLS^®^, doctors in over 80 countries have been
trained in PTC, and the programme has been endorsed by the World Health
Organization (WHO), among other international organizations
including several professional societies [@ptcfPtc2018].

There are observational studies indicating that trauma life support
training programmes may improve patient outcomes [@Jin2021], but there
are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020]. 
We performed a pilot study aiming to assess the feasibility of
conducting a cluster randomised controlled trial comparing ATLS and
PTC with standard care.  The objectives of this pilot study were
to:

- Estimate probable effect sizes on patient outcomes associated with
  ATLS^®^ and PTC compared with standard care, estimate the proportion of
  participants with the outcome (if binary), and estimate the
  intracluster correlation coefficient, as a basis for future sample
  size calculations.

@@

## Trial Design
We piloted a three-armed cluster randomised controlled trial. There
were two intervention arms, ATLS^®^ and PTC training, and one control
arm, standard care. We collected data for four months in all three
arms, first during a one month observation phase and then during a
three month intervention phase (or continued observation in the
control arm). This design allowed us to assess outcomes both as
final values and as change from baseline.

@@

## Study Setting
We conducted this pilot in seven Indian tertiary hospitals, where
neither ATLS^®^, PTC, nor any other trauma life support training program
is routinely taught. The seven hospitals were Seth G. S. Medical
College and K.E.M. Hospital, Mumbai; Lokmanya Tilak Municipal Medical
College and General Hospital, Mumbai; HBT Medical College And Dr. R N
Cooper Municipal General Hospital, Mumbai; Medica Superspecialty
Hospital, Kolkata; Medical College Kolkata, Kolkata; Sir Nil Ratan
Sircar Medical College & Hospital, Kolkata; Postgraduate Institute of
Medical Education & Research, Chandigarh.

@@

### Resident Doctor Participants
Resident doctors doing their speciality training in surgery or
emergency medicine managing trauma patients in the emergency
department and who were expected to remain in the participating
hospitals for at least one year. To facilitate administration each
surgical department is divided into units, which manage the out-patient 
and emergency departments, operating rooms etc on
different days each week. One or two, out of typically six, units'
residents were selected from each hospital. One unit consists of at
least three faculty and three to twelve residents.

To be eligible, units had to have a maximum of 25% of the doctors
trained in either ATLS^®^, PTC, or similar training programs before the
start of the pilot. Those residents who had received training in the
last five years were considered as trained. The figure of 25% was
decided through consensus in the research team, to balance feasibility
and contamination of results.

Consent was sought from the residents in each of the intervention
groups before they underwent the ATLS^®^ or PTC training. The consent
form for residents is available as Supplemental Material 2. We did
not ask for consent from residents at the units in the control
hospitals as their practice was not affected by this pilot and we
did not collect any personal identifiable data on them.

@@

## Interventions
In each intervention arm one or two units', out of typically six,
residents per hospital providing emergency care to trauma patients
were trained in either ATLS^®^ or PTC. For the purpose of this pilot
study, our target was to train a minimum of 75% of residents in each
unit. If residents dropped out or changed units after training but
before data collection was completed we planned to conduct additional
training if needed to meet the 75% criterion, but this was not
required. We did not train the units' faculty, as they are typically
not directly involved in the initial management of trauma patients.

The ATLS^®^ training was conducted in an ATLS certified training centre
in Mumbai, according to the standard ATLS^®^ curriculum
[@acsAtls2018]. The PTC training was conducted in New Delhi, according
to the standard PTC curriculum [@ptcfPtc2018]. These courses were
conducted over a period of 2.5 to 3 days. The residents certified
"pass" were considered as trained in respective courses.

The control group provided standard care with no intervention.

### Modifications
Both ATLS^®^ and PTC are standard training programs with fixed curricula
[@acsAtls2018; @ptcfPtc2018]. We did not modify the delivery or
content of these programs during this pilot.

### Adherence
The intervention was the training in either ATLS^®^ or PTC and resident
participants were required to adhere to, i.e. participate in, the
training, to be eligible for passing. We did not consider adherence to
training contents during care delivery as adherence to the trial
intervention.

@@

Although training in an ATLS^®^ is part of the curriculum it is 
optional and not doing this training does not
result in failure to obtain post graduation completion.

@@

The primary participant outcome was all cause mortality within 30 days
from the time of arrival to the emergency department. The primary
outcome and most secondary outcome were assessed and compared both as
final values and as change from baseline. All outcomes that pertain to
the individual participant level are detailed in Supplemental Material
3. We decided to include a large number of outcomes,
including some more exploratory, so that we could test their feasibility
and relevance.

@@

### Patients
Patients were screened for eligibility as they arrived at the emergency
department. Eligible patients were approached in the emergency
department to consent to follow up, if they were conscious. If they
were unconscious a patient representative was approached to consent to
follow up. Once the patient was conscious we approached the patient to
affirm the patient representative's consent. We followed up patients
at discharge, at 24 hours after arrival at the emergency department,
and at 30 days after arrival at the emergency department.

### Residents
Participating units were screened for eligibility once hospitals
confirmed their participation. All residents in eligible units were
approached to consent to training if their hospital was randomised to
either of the intervention arms. Training was conducted approximately
one month after the study started.

@@

## Patient and public involvement
In this study, we conducted community consultations to collect
inputs from patients, their caregivers, patient groups, and resident
doctors to be used in the selection of outcome measures and
implementation of the full-scale trial; following the Guidance for
Reporting Involvement of Patients and the Public (GRIIPS 2)
[@Staniszewska2017]. The results of these consultations will be reported 
separately.

## Data management
We used the online data collection tool KoboToolbox [@KoboToolbox],
accessible only over a virtual private network (VPN), for each
participating hospital to upload pseudonymised data to secure
servers. Data validation techniques such as restricted values or
values within a specific range were used to avoid ambiguous data
entries and ensure the validity of the data. Ambiguous responses and
errors were resolved after discussion with the core team during weekly
meetings. An instruction manual or codebook for data variables was
prepared to ensure consistency in data entry. This manual was referred
to during the project data collection and variable descriptions were
visible for each variable in the online data collection
tool. Pseudonymised data were stored at the centralised server. The
data were only accessible by the project's principal investigator or
by delegation by the project principal investigator.

@@

# Competing Interests
Several authors are ATLS^®^ and/or PTC instructors.

@@

```{r secondary-outcomes, echo = FALSE, results = "asis"}
table1::t1kable(secondary.outcomes.table, longtable = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8)
```
